Catecholamines and Hypoglycemia

### 2024-09-18 Christine Chan/EnVision CF/catecholamines_analysis.qmd
envision catecholamine update

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(gtsummary)
library(lmerTest)
library(emmeans)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r data cleaning}
# Import cleaned data
df <- read.csv("./Christine Chan/EnVision CF/Data_Clean/analysis_dataset.csv",
  na.strings = ""
)
# Timepoint and variables lists
timepoints <- c(0, 10, 30, 60, 90, 120, 150, 180)
limited_timepoints <- c(0, 120, 150, 180)
glucoses <- paste0("Glucose_", timepoints)
insulins <- paste0("Insulin_", timepoints)
epinephrines <- paste0("Epinephrine_", timepoints)
norepinephrines <- paste0("Norepinephrine_", timepoints)
# For this analysis, we only want those with catecholamines measured
df$catecholamines <- apply(
  df[, c(epinephrines, norepinephrines)], 1,
  function(r) {
    any(!is.na(r))
  }
)
df <- df %>% filter(catecholamines)
# Make a long dataset for mixed models
lmm_df <- df %>%
  select(
    study_id, Hypo60, Hypo70,
    all_of(c(glucoses, insulins, epinephrines, norepinephrines))
  ) %>%
  pivot_longer(all_of(c(glucoses, insulins, epinephrines, norepinephrines)),
    names_to = c(".value", "Timepoint"),
    names_pattern = "(.*)_(.*)"
  )
```

@@

```{r}
#| label: tbl-1
#| tbl-cap: Participant characteristics (overall)
df %>%
  select(
    redcap_data_access_group, age_visit:ethnicity, cftr_mutation_1,
    cftr_mutation_2
  ) %>%
  tbl_summary(by = redcap_data_access_group, missing_text = "Missing") %>%
  add_overall()
```

@@

```{r}
#| label: tbl-70
#| tbl-cap: BG < 70 at Any Point
#| warning: false
df %>%
  select(
    Hypo70, age_visit:cftr_mutation_2, hypoglycemia_symptoms_yesno,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo70, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

@@

```{r}
#| label: tbl-60
#| tbl-cap: BG < 60 at Any Point
#| warning: false
df %>%
  select(
    Hypo60, age_visit:cftr_mutation_2, hypoglycemia_symptoms_yesno,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo60, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

>>>>

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(gtsummary)
library(lmerTest)
library(emmeans)
library(GLMMadaptive)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r data cleaning}
# Import cleaned data
df <- read.csv("./Christine Chan/EnVision CF/Data_Clean/analysis_dataset.csv",
  na.strings = ""
)
# Timepoint and variables lists
timepoints <- c(0, 10, 30, 60, 90, 120, 150, 180)
limited_timepoints <- c(0, 120, 150, 180)
glucoses <- paste0("Glucose_", timepoints)
insulins <- paste0("Insulin_", timepoints)
epinephrines <- paste0("Epinephrine_", timepoints)
norepinephrines <- paste0("Norepinephrine_", timepoints)
adren_survey <- paste0("adren_score_", limited_timepoints)
neuro_survey <- paste0("neuro_score_", limited_timepoints)
total_survey <- paste0("total_score_", limited_timepoints)
num_symptoms_survey <- paste0("num_symptoms_", limited_timepoints)
hypo_surveys <- c(adren_survey, neuro_survey, total_survey, num_symptoms_survey)
# For this analysis, we only want those with catecholamines measured
df$catecholamines <- apply(
  df[, c(epinephrines, norepinephrines)], 1,
  function(r) {
    any(!is.na(r))
  }
)
df <- df %>% filter(catecholamines)
# Make a long dataset for mixed models
lmm_df <- df %>%
  select(
    study_id, Hypo60, Hypo70,
    all_of(c(glucoses, insulins, epinephrines, norepinephrines, hypo_surveys))
  ) %>%
  pivot_longer(
    all_of(c(glucoses, insulins, epinephrines, norepinephrines, hypo_surveys)),
    names_to = c(".value", "Timepoint"),
    names_pattern = "(.*)_([0-9]{1,3})"
  )
lmm_df$Timepoint <- factor(lmm_df$Timepoint, levels = timepoints)
# Remove BMI percentile for those >= 20
df$bmi_perc[df$age_visit >= 20] <- NA
df$bmi[df$age_visit < 20] <- NA
```

@@

```{r}
#| label: tbl-1
#| tbl-cap: Participant characteristics (overall)
df %>%
  select(
    redcap_data_access_group, age_visit:ethnicity, bmi, bmi_perc, CFTR
  ) %>%
  tbl_summary(by = redcap_data_access_group, missing_text = "Missing") %>%
  add_overall()
```

@@

```{r}
#| label: tbl-70
#| tbl-cap: BG < 70 at Any Point
#| warning: false
df %>%
  select(
    Hypo70, age_visit:ethnicity, bmi, bmi_perc, CFTR,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo70, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

## Hypoglycemia symptoms

### `adren_score`

`adren_score` was modeled using a generalized linear mixed effect model assuming a Poisson distribution for the outcome and a log link function. We also considered a negative binomial model, but the Poisson model was better by AIC. AICs for the Poisson and a zero-inflated Poisson were similar, so the Poisson was chosen for ease of interpretation. The following models were also determined based on AIC.

```{r}
#| warning: false
# Fit
# Poisson model better than NB by AIC and BIC
mod <- glmer(adren_score ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df,
  family = "poisson"
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates adren_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `neuro_score`

`neuro_score` was modeled using a zero-inflated Poisson model with random intercept for participant and an intercept as the only fixed effect in the zero part of the model.

```{r}
#| warning: false
# Fit
# Poisson model no different from NB by AIC
t <- lmm_df %>% filter(Timepoint %in% limited_timepoints)
mod <- mixed_model(neuro_score ~ Timepoint * Hypo70,
  random = ~ 1 | study_id, data = t,
  family = zi.poisson(), zi_fixed = ~1
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates neuro_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `num_symptoms`

`num_symptoms` was modeled using a Poisson model with random intercept for participant.

```{r}
#| warning: false
# Fit
# Poisson better than NB by AIC
mod <- glmer(num_symptoms ~ Timepoint * Hypo70 + (1 | study_id),
  data = lmm_df,
  family = "poisson"
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates num_symptoms") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `total_score`

`total_score` was modeled using a zero-inflated Poisson model with random intercept for participant and an intercept and hypoglycemia group as the fixed effects in the zero part of the model.

```{r}
#| warning: false
# Fit
mod <- mixed_model(total_score ~ Timepoint * Hypo70,
  random = ~ 1 | study_id, data = t,
  family = zi.poisson(), zi_fixed = ~Hypo70
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates total_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
#| label: tbl-60
#| tbl-cap: BG < 60 at Any Point
#| warning: false
df %>%
  select(
    Hypo60, age_visit:ethnicity, bmi, bmi_perc, CFTR,
    all_of(paste0("Glucose_", timepoints)),
    all_of(paste0("Insulin_", timepoints)),
    all_of(paste0("Epinephrine_", limited_timepoints)),
    all_of(paste0("Norepinephrine_", limited_timepoints))
  ) %>%
  tbl_summary(by = Hypo60, missing_text = "Missing") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

## Hypoglycemia symptoms

### `adren_score`

`adren_score` was modeled using a generalized linear mixed effect model assuming a Poisson distribution for the outcome and a log link function. We also considered a negative binomial model, but the Poisson model was better by AIC. AICs for the Poisson and a zero-inflated Poisson were similar, so the Poisson was chosen for ease of interpretation. The following models were also determined based on AIC.

```{r}
#| warning: false
# Fit
# Poisson model better than NB by AIC and BIC
mod <- glmer(adren_score ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df,
  family = "poisson"
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates adren_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `neuro_score`

`neuro_score` was modeled using a zero-inflated Poisson model with random intercept for participant and an intercept as the only fixed effect in the zero part of the model.

```{r}
#| warning: false
# Fit
# Poisson model no different from NB by AIC
t <- lmm_df %>% filter(Timepoint %in% limited_timepoints)
mod <- mixed_model(neuro_score ~ Timepoint * Hypo60,
  random = ~ 1 | study_id, data = t,
  family = zi.poisson(), zi_fixed = ~1
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates neuro_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `num_symptoms`

`num_symptoms` was modeled using a Poisson model with random intercept for participant.

```{r}
#| warning: false
# Fit
# Poisson better than NB by AIC
mod <- glmer(num_symptoms ~ Timepoint * Hypo60 + (1 | study_id),
  data = lmm_df,
  family = "poisson"
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates num_symptoms") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```

### `total_score`

`total_score` was modeled using a zero-inflated Poisson model with random intercept for participant and an intercept and hypoglycemia group as the fixed effects in the zero part of the model.

```{r}
#| warning: false
# Fit
mod <- mixed_model(total_score ~ Timepoint * Hypo60,
  random = ~ 1 | study_id, data = t,
  family = zi.poisson(), zi_fixed = ~Hypo60
)
mod %>% tbl_regression(intercept = T, exponentiate = T)
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"), type = "response"))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Rate", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
means %>%
  gt(caption = "LS Rates") %>%
  fmt_number(decimals = 2)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Rate, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = limited_timepoints) +
  ylab("LS Rates total_score") +
  xlab("Timepoint") +
  theme_bw() +
  theme(legend.title = element_blank())
```
